---
title: "behavioral analysis"
author: "Linda Sempf"
output:
  html_document:
    theme: sandstone
    highlight: rstudio
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: yes
    toc_depth: 3
  pdf_document:
    toc: yes
---


```{r setup, message=FALSE, warning=FALSE, comment=FALSE, echo=FALSE}
# rm(list = ls())
library(tidyverse); library(afex); library(emmeans); library('plot.matrix'); library(dplyr); library(readr); library(viridis); library(lme4); library(cluster); library(patchwork); library(philentropy); library(ggbeeswarm);
library(tibble); library(knitr); library(kableExtra); library(ggbreak);
```

which subjects 
```{r}
subjects_cuetarget <- c(seq(from="009",to="24"),seq(from="027",to="047"),seq(from="049",to="051"), seq(from="053",to="054")) 
subjects_cuetarget_task      <- setdiff(subjects_cuetarget, c(10,18,21,36,49,50,51)) 

subjects_detek <- c(seq(from="001",to="036")) 
subjects_detek <- setdiff(subjects_detek,c(32)) 

subjects_diract <- c(seq(from="001",to="018")) 
subjects_diract <- setdiff(subjects_diract, c(3,17)) 
```

# load data 
trained subjects
```{r, warning = FALSE, message = FALSE, echo=FALSE, results = FALSE}
# load data 
pathToTSVs = "data_trained"

loadSubject <- function(subject,df = data.frame()){
  for (session in 4){
    for (iRun in seq(1,4)){
      filename = sprintf('%s/sub-%03i/sub-%03i_ses-%03i_task-active_run-%i_events.tsv',
                         pathToTSVs, subject, subject, session, iRun)
      if (file.exists(filename)) {
        df_tmp <- read_tsv(filename,
                           col_types = cols(
                             onset = col_double(),
                             duration = col_double(),
                             trial_type = col_factor(),
                             response_time = col_character() # import response time as character 
                             # response_time = col_double()  # original code 
                           )
        ) %>%
          # append info which was part of filename
          mutate(subject = subject,
                 session = session,
                 run     = iRun
          )
        
        df <- rbind(df,df_tmp)     
      } else {
        print(paste("file", filename, "not found. moving on.."))
      }
    }
  }
  return(df)
}
l <- lapply(subjects_cuetarget_task, loadSubject)
df_cuetarget <- do.call("rbind", l) # dataframe with all subjects, all active sessions, all runs 

df_cuetarget$response_time <- as.numeric(df_cuetarget$response_time) # response_time = character; NAs introduced by coercion 
df_cuetarget[is.na(df_cuetarget)] <- NaN 

df_cuetarget <- df_cuetarget %>%
  mutate(delay = fct_collapse(factor(trial_type), 
                              short = c("shortExpected", "shortUnexpected"),
                              long = c("longExpected","longUnexpected")),
         expectation = fct_collapse(factor(trial_type),
                                    expected = c("shortExpected","longExpected"),
                                    unexpected = c("shortUnexpected","longUnexpected")
         )
  ) %>%
  mutate(subject = factor(subject))

# recalibration (beamer hat puffer: stimulus kam eigentlich später als bei PTB angezeigt)
lag_projector <- 0.135 # in seconds, like RTs 
df_cuetarget <- df_cuetarget %>%
  mutate(response_time = ifelse(session == 4, response_time-lag_projector, response_time))

# add 1:100 for each run
numel = nrow(df_cuetarget)/100 
iTrial = rep(c(1:100), times=numel)
df_cuetarget$iTrial = iTrial

# add 1:50 for each block
numel = (nrow(df_cuetarget)/100)*2 
iTrialBlock = rep(c(1:50), times=numel)
df_cuetarget$iTrialBlock = iTrialBlock

df_cuetarget$experiment = "cuetarget"

too_quick = 0.150 

upper_bound <- quantile(df_cuetarget$response_time, 0.99, na.rm = TRUE)
too_slow = upper_bound

df_cuetarget$subject_unique <- paste(df_cuetarget$experiment, df_cuetarget$subject, sep = "_")

df_cuetarget$training_status <- "trained"
df_cuetarget$lab <- "MRI"
```

load naive behavioral lab 
```{r, warning = FALSE, message = FALSE, echo=FALSE, results = FALSE}
pathToTSVs = "data_naive_behavioral-lab"

# create a function (for subjects with all measuremnets)
loadSubject <- function(subject,df = data.frame()){
  for (session in seq(1)){  # with detection task
    for (task in c("active")) {
      for (iRun in seq(1,4)){
        filename = sprintf('%s/sub-%03i/sub-%03i_ses-%03i_task-%s_run-%i_events.tsv',
                           pathToTSVs, subject, subject, session,task, iRun)
        if (file.exists(filename)) {
          df_tmp <- read_tsv(filename,
                            col_types = cols(
                              onset = col_double(),
                              duration = col_double(),
                              trial_type = col_factor(),
                              response_time = col_character() # import response time as character 
                            )
          ) %>%
            # append info which was part of filename
            mutate(subject = subject,
                  session = session,
                  run     = iRun
            )
        
          df <- rbind(df,df_tmp)     
        } else {
          print(paste("file", filename, "not found. moving on.."))
        }
      }
    }
  }
    return(df)
}

# and create data frame 
l <- lapply(subjects_detek, loadSubject)
df_detek <- do.call("rbind", l) # dataframe with all subjects, all active sessions, all runs 

df_detek$response_time <- as.numeric(df_detek$response_time) # response_time = character; NAs introduced by coercion 
df_detek[is.na(df_detek)] <- NaN 

# add columns 
df_detek <- df_detek %>%
  mutate(delay = fct_collapse(factor(trial_type), 
                              short = c("shortExpected", "shortUnexpected"),
                              long = c("longExpected","longUnexpected")),
         expectation = fct_collapse(factor(trial_type),
                                    expected = c("shortExpected","longExpected"),
                                    unexpected = c("shortUnexpected","longUnexpected")
         )
  ) %>%
  mutate(subject = factor(subject))

# add 1:104 for each run
numel = nrow(df_detek)/100
iTrial = rep(c(1:100), times=numel)
df_detek$iTrial = iTrial

# add 1:52 for each block
numel = (nrow(df_detek)/100)*2 
iTrialBlock = rep(c(1:50), times=numel) 
df_detek$iTrialBlock = iTrialBlock

df_detek$experiment = "detek"

df_detek$subject_unique <- paste(df_detek$experiment, df_detek$subject, sep = "_")

df_detek$training_status <- "naive"

df_detek$lab <- "beh"
```

load naive MRI
```{r, warning = FALSE, message = FALSE, echo=FALSE, results = FALSE}
pathToTSVs = "data_naive_MRI-lab"

# create a function (for subjects with all measuremnets)
loadSubject <- function(subject,df = data.frame()){
  for (session in seq(1)){  # with detection task
    for (task in c("active")) {
      for (iRun in seq(1,4)){
        filename = sprintf('%s/sub-%03i/sub-%03i_ses-%03i_task-%s_run-%i_events.tsv',
                           pathToTSVs, subject, subject, session,task, iRun)
        if (file.exists(filename)) {
          df_tmp <- read_tsv(filename,
                            col_types = cols(
                              onset = col_double(),
                              duration = col_double(),
                              trial_type = col_factor(),
                              response_time = col_character() # import response time as character 
                            )
          ) %>%
            # append info which was part of filename
            mutate(subject = subject,
                  session = session,
                  run     = iRun
            )
        
          df <- rbind(df,df_tmp)     
        } else {
          print(paste("file", filename, "not found. moving on.."))
        }
      }
    }
  }
    return(df)
}

# and create data frame 
l <- lapply(subjects_diract, loadSubject)
df_diract <- do.call("rbind", l) # dataframe with all subjects, all active sessions, all runs 

df_diract$response_time <- as.numeric(df_diract$response_time) # response_time = character; NAs introduced by coercion 
df_diract[is.na(df_diract)] <- NaN 

# add columns 
df_diract <- df_diract %>%
  mutate(delay = fct_collapse(factor(trial_type), 
                              short = c("shortExpected", "shortUnexpected"),
                              long = c("longExpected","longUnexpected")),
         expectation = fct_collapse(factor(trial_type),
                                    expected = c("shortExpected","longExpected"),
                                    unexpected = c("shortUnexpected","longUnexpected")
         )
  ) %>%
  mutate(subject = factor(subject))

# recalibration (beamer hat puffer: stimulus kam eigentlich später als bei PTB angezeigt)
lag_projector <- 0.135 # in seconds, like RTs 
df_diract <- df_diract %>%
  mutate(response_time = ifelse(session == 1, response_time-lag_projector, response_time))

# add 1:104 for each run
numel = nrow(df_diract)/100
iTrial = rep(c(1:100), times=numel)
df_diract$iTrial = iTrial

# add 1:52 for each block
numel = (nrow(df_diract)/100)*2 
iTrialBlock = rep(c(1:50), times=numel) 
df_diract$iTrialBlock = iTrialBlock

df_diract$experiment = "diract"

df_diract$subject_unique <- paste(df_diract$experiment, df_diract$subject, sep = "_")

# df_diract$response_time <- df_diract$response_time - 0.125

df_diract$training_status <- "naive"

df_diract$lab <- "MRI"
```

combine data frames
```{r}
df <- rbind(df_cuetarget, df_detek, df_diract)
rm(df_cuetarget,df_detek,df_diract,l)

df$subject_unique <- as.factor(df$subject_unique)
```


# TE effect in trained participants 
t test: responses early likely VS. early unlikely
```{r}
fit_TE_cuetarget <- aov_ez(
  id = "subject_unique",
  dv = "response_time",
  within = c("expectation"),
  data = df %>% subset(experiment == "cuetarget") %>% 
    subset(response_time < 1) %>% # drop too slow errors
    subset(delay=="short"), 
  fun_aggregate = mean
)
fit_TE_cuetarget

means_early <- emmeans(fit_TE_cuetarget, ~ expectation)
pairs(means_early)
```


# table 1: response times  
session-wise
```{r}
df %>%
  subset(response_time<1) %>%
  filter(delay == "short") %>%
  group_by(experiment, subject_unique, session, delay, expectation) %>%
  summarize(mean_RT = mean(response_time, na.rm = TRUE), .groups = "drop") %>%
  group_by(experiment, session, delay, expectation) %>%
  summarize(
    mean_RT_group = mean(mean_RT, na.rm = TRUE),
    se = sd(mean_RT, na.rm = TRUE) / sqrt(n()),  
    n = n(),                                    
    .groups = "drop" ) %>%
  mutate(mean_RT_group = mean_RT_group * 1000,
         se = se * 1000)
```

run-wise
```{r}
table_RT <- df %>%
  subset(response_time<1) %>%
  filter(delay == "short") %>%
  group_by(experiment, subject_unique, session, run, delay, expectation) %>%
  summarize(mean_RT = mean(response_time, na.rm = TRUE), .groups = "drop") %>%
  group_by(experiment, session, run, delay, expectation) %>%
  summarize(
    mean_RT_group = mean(mean_RT, na.rm = TRUE),
    se = sd(mean_RT, na.rm = TRUE) / sqrt(n()),  
    n = n(),                                    
    .groups = "drop" ) %>%
  mutate(mean_RT_group = mean_RT_group * 1000,
         se = se * 1000) %>%
  mutate(
    mean_RT_group = round(mean_RT_group, 2),
    se = round(se, 2)) %>%
  select(experiment, session, run, expectation, mean_RT_group, se) %>%
  mutate(
    expectation = dplyr::recode(
      expectation,
      "likely"   = "early likely",
      "unlikely" = "early unlikely"
    ),
    experiment = dplyr::recode(
      experiment,
      "cuetarget" = "fMRI experiment",
      "detek"     = "behavioral experiment",
      "diract"    = "fMRI experiment: ohne training"
    ),
    session = dplyr::recode(
      session,
      "1" = "1 active naive",
      "2" = "2 active training",
      "3" = "3 active training",
      "4" = "4 active trained"
    )
  ) %>%
  mutate(mean_RT_group = round(mean_RT_group, 2),
         se = round(se, 2)) %>%
  pivot_wider(names_from  = expectation,
    values_from = c(mean_RT_group, se),
    names_glue  = "{expectation}_{.value}") %>%
  relocate(
    experiment, session, run,
    `expected_mean_RT_group`, `expected_se`,
    `unexpected_mean_RT_group`, `unexpected_se`) %>%
  rename(
    "early likely M"    = "expected_mean_RT_group",
    "early likely SE"   = "expected_se",
    "early unlikely M"  = "unexpected_mean_RT_group",
    "early unlikely SE" =  "unexpected_se"
  ) %>%
  mutate(dplyr::across(
      c(`early likely M`, `early likely SE`,
        `early unlikely M`, `early unlikely SE`),
      ~ sprintf("%.2f", .x)))
table_RT
```


# ANOVA in trained participants 
trained participants, early trials 
```{r}
fit_TE_cuetarget_early <- aov_ez(
  id = "subject_unique",
  dv = "response_time",
  within = c("run", "expectation"),
  data = df %>% subset(experiment == "cuetarget") %>% 
    subset(response_time < 1) %>% # drop too slow errors
    subset(delay=="short"), 
  fun_aggregate = mean
)
fit_TE_cuetarget_early
```

trained participants, late trials 
```{r}
fit_TE_cuetarget_late <- aov_ez(
  id = "subject_unique",
  dv = "response_time",
  within = c("run", "expectation"),
  data = df %>% subset(experiment == "cuetarget") %>% 
    subset(response_time < 1) %>% # drop too slow errors
    subset(delay=="long"), 
  fun_aggregate = mean
)
fit_TE_cuetarget_late
```


# figure 2: response times 
```{r}
text_size <- 10
text_face <- "plain"

df_rt <- df %>%
  subset(response_time < 1) %>% # drop too slow errors
  subset(experiment=="cuetarget") %>%
  subset(session==4) %>%
  group_by(experiment, subject_unique, session, delay, expectation) %>%
  summarize(mean_RT = mean(response_time, na.rm = TRUE), .groups = "drop") %>%
  filter(!(is.na(mean_RT))) %>%
  mutate(
    session_num = as.integer(as.character(session)),
    timeline    = if_else(experiment == "detek", session_num - 1L, session_num)
  ) %>%
  filter(timeline >= 0)

sum_rt <- df_rt %>%
  group_by(timeline, experiment, delay, expectation) %>%
  summarize(
    mean_group = mean(mean_RT, na.rm = TRUE),
    se_group   = sd(mean_RT, na.rm = TRUE) / sqrt(dplyr::n()),
    .groups    = "drop"
  ) %>%
  mutate(
    exp_group = paste(delay, expectation),
    exp_group = factor(
      exp_group,
      levels = c("short expected","short unexpected",
                 "long expected","long unexpected")
    ))

label_map <- c(
  `0` = "active naive (session 1)",
  `2` = "active training (session 2)",
  `3` = "active training (session 3)",
  `4` = "active trained (session 4)")

sum_rt <- sum_rt %>%
  mutate(timeline = factor(timeline, levels = c(0, 2, 3,4), ordered = TRUE))


pd <- position_dodge(width = 0.6)


# RT distribution plots (session 4, both delays)
p2 <- df %>%
  subset(experiment=="cuetarget") %>%
  subset(session==4) %>%
  subset(response_time < 1) %>% # drop too slow errors
  group_by(experiment, subject_unique, session, delay, expectation) %>% # aggregate data per subject
  summarize(mean_RT = mean(response_time, na.rm =T),
            median_RT = median(response_time, na.rm =T)) %>%
  mutate(
    condition = paste(delay, expectation), condition = recode(
      condition, "short expected"   = "early likely", "short unexpected" = "early unlikely",
      "long expected"    = "late likely", "long unexpected"  = "late unlikely"),
    condition = factor(condition, levels = c("early likely","early unlikely","late likely","late unlikely"))) %>%
  subset(!is.na(mean_RT)) %>%
  ggplot(aes(x = mean_RT, color = condition, xName='weight')) +
  geom_density(size=2) +
  #geom_density(aes(y=..scaled..), size=2) +
  scale_color_manual(values = c("late likely"="#756bb1","late unlikely"="#74c476",
                                "early likely"="#54278f","early unlikely"="#238b45")) +
  ggtitle(paste0("")) +
  xlab("mean response time") +
  ylab("Probability Density in au") +
  xlim(c(-0.20,0.4)) +
  theme_minimal() +
  theme(legend.title = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.y = element_text(size = 8))

p1_base <- ggplot(sum_rt, aes(x = experiment, y = mean_group, fill = exp_group)) +
  facet_wrap(
    ~ timeline, nrow = 1, scales = "free_x",
    labeller = labeller(timeline = function(x) label_map[as.character(x)])
  ) +
  geom_col(position = pd, width = 0.5) +
  geom_errorbar(
    aes(ymin = mean_group - se_group, ymax = mean_group + se_group),
    position = pd, width = 0.15, linewidth = 1.0
  ) +
  scale_fill_manual(values = c(
    "long expected"="#756bb1","long unexpected"="#74c476",
    "short expected"="#54278f","short unexpected"="#238b45"
  ), drop = FALSE) +
  labs(x = NULL, y = "mean RT (s)", title = "") +
  theme_minimal(base_size = text_size) +
  theme(
    text = element_text(face = text_face),
    panel.grid.minor = element_blank(),
    legend.position = "none",
    strip.text = element_text(face = text_face),
    axis.title.y = element_text(size = 8)
  )

p1_top <- p1_base +
  coord_cartesian(ylim = c(0.145, 0.29)) +
  scale_y_continuous(
    breaks = c(0.135, 0.160, 0.185, 0.210, 0.235, 0.260)
  ) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank(),
    plot.margin = margin(5.5, 5.5, 0, 5.5)
  )

p1_bottom <- p1_base +
  coord_cartesian(ylim = c(0, 0.0008)) +
  scale_y_continuous(
    breaks = 0,
    labels = "0"
  ) +
  labs(title = NULL, y = NULL) +
  theme(
    strip.text = element_blank(),
    plot.margin = margin(0, 5.5, 5.5, 5.5)
  )

p1_broken <- p1_top / p1_bottom +
  plot_layout(heights = c(80, 1))

(p1_broken | p2) +
  plot_layout(widths = c(1, 2))
```


# ANOVA in naive participants 
naive participants, early trials 
```{r}
fit_TE_naive_early <- aov_ez(
  id = "subject_unique",
  dv = "response_time",
  within = c("run", "expectation"),
  data = df %>% subset(training_status == "naive") %>% 
    subset(response_time < 1) %>% # drop too slow errors
    subset(delay=="short"), 
  fun_aggregate = mean
)
fit_TE_naive_early
```

naive participants, late trials
```{r}
fit_TE_naive_late <- aov_ez(
  id = "subject_unique",
  dv = "response_time",
  within = c("run", "expectation"),
  data = df %>% subset(training_status == "naive") %>% 
    subset(response_time < 1) %>% # drop too slow errors
    subset(delay=="long"), 
  fun_aggregate = mean
)
fit_TE_naive_late
```


# trained VS. naive participants 
linear mixed-effects model, early trials 
```{r}
fit <- mixed(
  response_time ~ expectation * training_status + lab +
    (1 | subject_unique),
  data = df %>%
    filter(
      delay == "short",
      response_time < 1,
      experiment %in% c("diract", "detek") |
      (experiment == "cuetarget" & session == 4)) %>%
    mutate(
      expectation = factor(expectation),
      training_status = factor(training_status),
      lab = factor(lab),
      subject_unique = factor(subject_unique)
    ),
  method = "S"
)
fit

emm <- emmeans(fit, ~ expectation * training_status)

contrast_expectation <- contrast(
  emm,
  interaction = "pairwise",
  by = NULL
)
contrast_expectation

# numeric TE effects 
df %>% 
  subset(delay == "short") %>%
  subset(experiment == "diract" | experiment == "detek" |
           (experiment == "cuetarget" & session == 4)) %>%
  subset(response_time < 1) %>%
  group_by(lab, training_status, experiment, subject_unique, session, run, delay, expectation) %>%
  summarize(mean_RT = mean(response_time, na.rm = TRUE)) %>%
  group_by(lab, training_status, experiment, subject_unique, session, delay, expectation) %>%
  summarize(mean_RT = mean(mean_RT)) %>%
  pivot_wider(names_from = expectation, values_from = mean_RT) %>%
  mutate(TE = unexpected - expected) %>%
  group_by(training_status) %>%
  summarize(meanTE = mean(TE)) %>%
  pivot_wider(
    names_from = training_status,
    values_from = meanTE) %>%
  mutate(difference_meanTE = naive - trained)
```

linear mixed-effects model, late trials 
```{r}
fit <- mixed(
  response_time ~ expectation * training_status + lab +
    (1 | subject_unique),
  data = df %>%
    filter(
      delay == "long",
      response_time < 1,
      experiment %in% c("diract", "detek") |
      (experiment == "cuetarget" & session == 4)) %>%
    mutate(
      expectation = factor(expectation),
      training_status = factor(training_status),
      lab = factor(lab),
      subject_unique = factor(subject_unique)
    ),
  method = "S"
)
fit
```